Medical Concept Extraction

### 2012-12-17 paper/paper.tex
Salman's final pass

This paper presents a system for extracting medical concepts from patient hospital records. Hospitals and medical practices are increasingly switching to electronic medical records. This trend has opened the doors to exciting applications for Natural Language Processing in medicine and biology. One such application is concept extraction: identifying high-level semantic labels in a body of text. Concept extraction on medical records is useful in automating analysis tasks that would otherwise have to be done manually. For example, hospital administrators may want to tabulate what medications are being prescribed the most. Insurance companies may want to ensure that patients receive proper care and are billed accordingly for fraud detection and auditing purposes. Government agencies may be interested in exploring and better understanding public health trends. Automatic concept extraction systems help to streamline such critical yet laborious tasks.

In service of that end, we present an algorithm and system that is capable of classifying words in medical records across four different medically-relevant categories: (1) problems, (2) tests, (3) treatments, and (4) non-relevant terms. This is useful not only in granting end users the ability to analyze medical records, but also in supporting other higher-level Natural Language Processing and learning systems. This paper describes the system's design, implementation, and performance when tested on set of 500 pre-annotated medical records from Boston-area hospitals.

@@

In this work, we present a system for extracting problems, tests, and treatments from medical records at comparable performance (approximately 0.8 F1 score on average in testing) to existing methods. Our classification strategies prove successful in spite our relatively limited data set, which is comprised of hospital records from two Boston area facilities. Our portable open-source system is implemented in python with easily configurable and extendable feature set and three classifier options.

An additional web-based service also allows users to input single sentences for concept extraction on the fly. With great ease of access, this system helps to advance a novel and important application of Natural Language Processing in the health care domain.

@@

We tested a non-linear radial basis kernel SVM as provided by the open-source library, LibSVM \cite{libsvm}.

@@

We additionally tested a strictly linear SVM using the open-source library, LibLinear \cite{liblinear}, which provides an L2-loss linear SVM solver.

@@

\subsubsection{Word}
The most obvious property of each word is perhaps the word itself. Since there is no linguistically meaningful, strict ordering over words, each word is represented as its own dimension. Consequently adding a baseline feature containing the word increases the dimension of our feature space to the size of the vocabulary in the training data. Nevertheless, the complete word serves as a fruitful feature in medical term classification in which many words carry the same meaning regardless of context. For example, terms such as "x-ray", "screen", "exam", "veal", and "sats" should be consistently recognized as test terms, just as "swelling", "failure", "trauma", and "syndrome" always indicate problems, and "therapy", "anesthesia", "vaccine", and "medication" are treatments. Granted, words like "heart" could be part of "heart rate" or "heart attack" and beg for more generic features.

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, as well as a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same type. Additional classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an many more dimensions to our feature space. By naively including all word shape alternatives as features, we were able to improve our toy test CRF F1 score from 0.82 to 0.85.

@@

\item {\tt Word Endings Indicative of Medical Problem Type}
Another obvious trend specific to medical problems are diseases ending in "itis" such as "bronchitis", "cellulites", "fibrositis", etc. and conditions ending in "ic" such as "diabetic", "anemic", etc. We added a regular expression to flag words with these endings, but the feature did little to nothing to improve performance. Such observations were likely already handled by our other word features.

\item {\tt Prognosis Locations}
Many diagnoses include locational information that should be tagged as part of the problem (i.e. "C4-C5 herniated disks"). We still think that a robust feature of this form could benefit our system, but our particular attempts proved unsuccessful.

\item {\tt Flag Words}
Certain terms such as those described in 5.2.1 are consistently labeled as "problems", "tests", or "treatments". We attempted to create small dictionaries of these terms and assign flags to words that fall into one of the three dictionaries. The issue with this strategy is that words that should be labeled as one of our three medical terms but do not appear in the dictionary are now more closely aligned with purely non-medical words. That said, this feature improved accuracy for our three categories, but also yielded far too many false positives.

@@

\subsubsection{Part of Speech}
A fruitful strategy for any concept extractor is to first tag words with labels indicating their part of speech. In our case, for example, whether an ambiguous term is used as an adjective versus a noun could distinguish it as part of a test rather than a problem. We used the NTLK max entropy Part of Speech Tagger, which is trained on the Penn TreeBank to tag each sentences. We then iterate over the returned tags and add these as word-level features, adding several dimensions to our feature set, one for each part of speech tag.

@@

\subsection{N-gram Features}
Again, although our primary goal is to classify individual words, a single word in a medical record entry is hardly independent from the next.  More specifically, medical concepts can often span more than one word. We may not be interested in detecting these exact spans, but features from certain words provide insight into how to label their neighbors. For example, ambiguous words such as "heart" could precede "rate" or "disease", and making that distinction between test and problem is much easier when we examine the subsequent word in the sentence containing "heart". A neighboring word can serve as an indicator even if it is not part of the same concept span. For example, words like "showed" and its synonyms are non-medical, but tend to follow a test term. Likewise, "rule out" and "history of" tend to precede problems. With this in mind, we observe all features of immediately neighboring words.

@@

\section{System}
The system we have developed consists of a framework in order to perform concept extraction, a web service frontend which allows users to test the service without seeing any of the underlying data, and the machine learning libraries that are leveraged.

@@

The first format is a straight forward tokenized text file that contains each sentence on a separate line. The test data for training the models were processed into this format. For cases where the sentence was not previously tokenized, the framework utilizes {\tt nltk}'s {\tt sent\_tokenize}  to break a string into sentences and then {\tt nltk}'s {\tt word\_tokenize} to break each sentence into normalized tokens.

@@

In addition to the various command line utilizes described in the previous section, our system also includes a Web-based service that allows users to manually input a single sentence, select a model to use for prediction, and see the predicted labels that model generated for the input. Figure~\ref{fig:web_interface} shows a screenshot of the web application in use.

@@

\item {\tt nltk} \cite{nltk}: nltk is a very popular NLP library written in Python. We leverage nltk to generate some of the features on our data. In particular, we utilize nltk's maximum-entropy part-of-speech tagging as well as many of its stemming algorithms.

\item {\tt flask} \cite{flask}: Flask is a micro Web framework written in Python. It contains a small yet useful set of features that allow developers to easily setup RESTful Web backends and an elegant templating system to gracefully generate HTML. Flask is used to build the Web service portion of the project.

@@

We deployed the system on a machine running Ubuntu 11.10 with a QuadCore AMD Opteron processor running a 2.5Ghz and 48GB of RAM.

@@

As priorly mentioned, we tested our feature set using three classifiers in search of the best concept extraction system (general SVM, linear SVM, and CRF), all of which yielded comparable results in terms of their average F1 scores, but showed significant difference in terms of efficiency. Our least time efficient classifier, SVM, was eventually eliminated from our system, as discussed in the subsequent section.

>>>>

This paper presents a system for extracting medical concepts from patient hospital records. Hospitals and medical practices are increasingly switching to electronic medical records. This trend has opened the doors to exciting applications for natural language processing in medicine and biology. One such application is concept extraction: identifying high-level semantic labels in a body of text. Concept extraction on medical records is useful in automating analysis tasks that would otherwise have to be done manually. For example, hospital administrators may want to tabulate what medications are being prescribed the most. Insurance companies may want to ensure that patients receive proper care and are billed accordingly for fraud detection and auditing purposes. Government agencies may be interested in exploring and better understanding public health trends. Automatic concept extraction systems help to streamline such critical yet laborious tasks.

In service of that end, we present an algorithm and system that is capable of classifying words in medical records across four different medically-relevant categories: (1) problems, (2) tests, (3) treatments, and (4) non-relevant terms. This is useful not only in granting end users the ability to analyze medical records, but also in supporting other higher-level natural language processing and learning systems. This paper describes the system's design, implementation, and performance when tested on set of 500 pre-annotated medical records from Boston-area hospitals.

@@

In this work, we present a system for extracting problems, tests, and treatments from medical records at comparable performance (approximately 0.8 F1 score on average in testing) to existing methods. Our classification strategies prove successful in spite our relatively limited data set, which is comprised of hospital records from two Boston area facilities. Our portable open-source system is implemented in Python with easily configurable and extendable feature set and three classifier options.

An additional Web-based service also allows users to input single sentences for concept extraction on the fly. With great ease of access, this system helps to advance a novel and important application of natural language processing in the health care domain.

@@

We tested a non-linear radial basis kernel SVM as provided by the open-source library, libsvm \cite{libsvm}.

@@

We additionally tested a strictly linear SVM using the open-source library, liblinear \cite{liblinear}, which provides an L2-loss linear SVM solver.

@@

\subsubsection{Word}
The most obvious property of each word is perhaps the word itself. Since there is no linguistically meaningful, strict ordering over words, each word is represented as its own dimension. Consequently adding a baseline feature containing the word increases the dimension of our feature space to the size of the vocabulary in the training data. Nevertheless, the complete word serves as a fruitful feature in medical term classification in which many words carry the same meaning regardless of context. For example, terms such as ``x-ray'', ``screen'',``exam'', ``veal'', and ``sats'' should be consistently recognized as test terms, just as ``swelling", ``failure", ``trauma", and ``syndrome" always indicate problems, and ``therapy", ``anesthesia", ``vaccine", and ``medication" are treatments. Granted, words like ``heart" could be part of ``heart rate" or ``heart attack" and beg for more generic features.

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein \cite{wordshape}, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, as well as a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same type. Additional classifiers developed by Christopher Manning \cite{wordshape} seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an many more dimensions to our feature space. By naively including all word shape alternatives as features, we were able to improve our toy test CRF F1 score from 0.82 to 0.85.

@@

\item {\tt Word Endings Indicative of Medical Problem Type}
Another obvious trend specific to medical problems are diseases ending in ``itis'' such as ``bronchitis'', ``cellulites'', ``fibrositis'', etc. and conditions ending in ``ic" such as ``diabetic", ``anemic", etc. We added a regular expression to flag words with these endings, but the feature did little to nothing to improve performance. Such observations were likely already handled by our other word features.

\item {\tt Prognosis Locations}
Many diagnoses include locational information that should be tagged as part of the problem (i.e. ``C4-C5 herniated disks"). We still think that a robust feature of this form could benefit our system, but our particular attempts proved unsuccessful.

\item {\tt Flag Words}
Certain terms such as those described in 5.2.1 are consistently labeled as ``problems", ``tests", or ``treatments". We attempted to create small dictionaries of these terms and assign flags to words that fall into one of the three dictionaries. The issue with this strategy is that words that should be labeled as one of our three medical terms but do not appear in the dictionary are now more closely aligned with purely non-medical words. That said, this feature improved accuracy for our three categories, but also yielded far too many false positives.

@@

\subsubsection{Part of Speech}
A fruitful strategy for any concept extractor is to first tag words with labels indicating their part of speech. In our case, for example, whether an ambiguous term is used as an adjective versus a noun could distinguish it as part of a test rather than a problem. We used the NTLK max entropy part of speech tagger, which is trained on the Penn TreeBank to tag each sentences. We then iterate over the returned tags and add these as word-level features, adding several dimensions to our feature set, one for each part of speech tag.

@@

\subsection{N-gram Features}
Again, although our primary goal is to classify individual words, a single word in a medical record entry is hardly independent from the next.  More specifically, medical concepts can often span more than one word. We may not be interested in detecting these exact spans, but features from certain words provide insight into how to label their neighbors. For example, ambiguous words such as ``heart" could precede ``rate" or ``disease", and making that distinction between test and problem is much easier when we examine the subsequent word in the sentence containing ``heart". A neighboring word can serve as an indicator even if it is not part of the same concept span. For example, words like "showed" and its synonyms are non-medical, but tend to follow a test term. Likewise, ``rule out" and ``history of" tend to precede problems. With this in mind, we observe all features of immediately neighboring words.

@@

\section{System}
The system we have developed consists of a framework in order to perform concept extraction, a Web service frontend which allows users to test the service without seeing any of the underlying data, and the machine learning libraries that are leveraged.

@@

The first format is a straight forward tokenized text file that contains each sentence on a separate line. The test data for training the models were processed into this format. For cases where the sentence was not previously tokenized, the framework utilizes {\tt NLTK}'s {\tt sent\_tokenize}  to break a string into sentences and then {\tt NLTK}'s {\tt word\_tokenize} to break each sentence into normalized tokens.

@@

In addition to the various command line utilizes described in the previous section, our system also includes a Web-based service that allows users to manually input a single sentence, select a model to use for prediction, and see the predicted labels that model generated for the input. Figure~\ref{fig:web_interface} shows a screenshot of the Web application in use.

@@

\item {\tt NLTK} \cite{nltk}: NLTK is a very popular NLP library written in Python. We leverage NLTK to generate some of the features on our data. In particular, we utilize NLTK's maximum-entropy part-of-speech tagging as well as many of its stemming algorithms.

\item {\tt Flask} \cite{flask}: Flask is a micro Web framework written in Python. It contains a small yet useful set of features that allow developers to easily setup RESTful Web backends and an elegant templating system to gracefully generate HTML. Flask is used to build the Web service portion of the project.

@@

We deployed the system on a machine running Ubuntu 11.10 with a QuadCore AMD Opteron processor running at 2.5Ghz and 48GB of RAM.

@@

As previously mentioned, we tested our feature set using three classifiers in search of the best concept extraction system (general SVM, linear SVM, and CRF), all of which yielded comparable results in terms of their average F1 scores, but showed significant difference in terms of efficiency. Our least time efficient classifier, SVM, was eventually eliminated from our system, as discussed in the subsequent section.
